---
title: "ps5"
author: "Xiangguang"
date: "2023-02-19"
output: html_document
---


  2. (2pt) Load data. How many rows/columns do we have?
```{r}
library(dplyr)
library(tidyverse)
```
```{r}
data <- read_delim("gapminder.csv")
dim(data)
```
    3. (2pt) Print a small sample of data. Does it look OK?
    
```{r}
rand_rows <- sample(nrow(data), 10)
data[rand_rows, ]

```
    
    
        2 Descriptive statistics (15pt)
  1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
  
```{r}
length(unique(data$iso2))
length(unique(data$name))
length(unique(data$iso3))
```

  2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
  even more iso3 -codes. What is going on? Can you find it out?
    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that
  correspond to more than one name? What are these countries?

  
    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
  more than one iso3-code? What are these countries?

    Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
countries <- unique(data[, c("name", "iso2", "iso3")])
countries
```
There are some valid iso3 names does not have a name and iso2 for it. 

```{r}
iso2names <- data %>% 
  group_by(iso2) %>% 
  summarise(num = n_distinct(name))
iso2names
```
There aren't any iso2 codes that correspond to more than one names. 
There are 
```{r}
iso3names <- data %>% 
  group_by(iso3) %>% 
  summarise(num = n_distinct(name))
iso3names
```
There aren't any iso3-code that have more than one name. 



  3. (2pt) What is the minimum and maximum year in these data?
  
```{r}
new_data <- data[complete.cases(data$time),]
max(new_data$time)
```
```{r}
min(new_data$time)
```
  3 CO2 emissions (30pt)
Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?

```{r}
sum(is.na(data$co2))
sum(is.na(data$co2_PC))
```

```{r}
missing <- data %>%
  group_by(time) %>%
  summarise(co2_missing = sum(is.na(data$co2)),
            co2_pc_missing = sum(is.na(data$co2_PC)))

missing_co2 <- missing[which.max(missing$co2_missing), "time"]
missing_co2_pc <- missing[which.max(missing$co2_pc_missing), "time"]

missing_co2

missing_co2_pc

```
For both co2 and co2_pc, 1960 has the most missing data.



2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.

```{r}
data %>% 
  filter(iso3 ==c( "USA", "CHN", "IND", "KOR", "SDN", "QAT")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2, col=iso3))+
  geom_point(aes(x=time, y = co2, col=iso3))
```

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?

```{r}
data %>% 
  filter(iso3 ==c( "USA", "CHN", "IND", "KOR", "SDN", "QAT")) %>% 
  ggplot()+
  geom_line(aes(x=time, y = co2_PC, col=iso3))+
  geom_point(aes(x=time, y = co2_PC, col=iso3))
```
QAT has a very high co2 emission per captita comparing to other countries.

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different
size.
```{r}
avg_co2_pc_continent_2016 <- data %>%
  filter(time == 2016) %>%
  group_by(region) %>%
  summarize(avg_co2_pc = mean(co2_PC, na.rm = TRUE))
avg_co2_pc_continent_2016$time <- 2016
avg_co2_pc_continent_2016 <- na.omit(avg_co2_pc_continent_2016)
avg_co2_pc_continent_2016
```
```{r}
avg_co2_pc_continent_1960 <- data %>%
  filter(time == 1960) %>%
  group_by(region) %>%
  summarize(avg_co2_pc = mean(co2_PC, na.rm = TRUE))
avg_co2_pc_continent_1960$time <- 1960
avg_co2_pc_continent_1960 <- na.omit(avg_co2_pc_continent_1960)
avg_co2_pc_continent_1960
```
In 2016, Europe has the highest average CO2_PC, and the second is Asia, the third is Americas. In 1960, Americas has the highest average CO2_PC, the second is Europe, and the third is Ocenia.
Hint: Americas 2016 should be 4.80.
5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r}
avg_co2_pc_both <- rbind(avg_co2_pc_continent_1960, avg_co2_pc_continent_2016)
ggplot(avg_co2_pc_both, aes(x = region, y = avg_co2_pc, fill = factor(time))) + 
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("purple", "orange")) +
  labs(x = "Continent", y = "Average CO2 (per capita)", fill = "Year")
```


Hint: it should look something along these lines:
  
  6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).

```{r}
valid_data <- data %>% filter(!is.na(co2_PC))
vliad_data_2016 <- valid_data %>% filter(time == 2016) %>%
  group_by(region) %>% 
  slice_max(order_by = co2_PC, n = 3) %>% 
  bind_rows(valid_data %>% filter(time == 2016) %>% 
              group_by(region) %>% 
              slice_min(order_by = co2_PC, n = 3))
vliad_data_2016
```
The largest 3: 
Africa：South Africa, Libya, Seychelles Americas: Trinidad and Tobago, USA, Canada Asia: Qatar, Kuwait, Bahrain Europe: Gibraltar, Luxembourg, Faroe Islands Oceania: New Caledonia, Australia, Palau

The smallest 3: 
Africa: Congo, Somalia, Burundi Americas: Haiti, Nicaragua, Honduras Asia: Afghanistan, Nepal, Yemen Europe: Liechtenstein, Albania, Republic of Moldova. Oceania: Solomon Islands, Vanuatu, Kiribati

4 GDP per capita (50pt)
Let’s look at GDP per capita (GDP_PC).
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r}
data_1960 <- data %>% filter(time == 1960)
ggplot(data_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point(alpha=0.7) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  scale_size_continuous(range = c(1, 10), breaks = c(1000000, 5000000, 10000000, 50000000, 100000000)) +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  labs(title = "GDP per capita versus life expectancy by country (1960)",
       subtitle = "Point size indicates total population, color indicates continent") +
  theme_bw()
```


2. (4pt) Make a similar plot, but this time use 2019 data only.
3

```{r}
data_2019 <- data %>% filter(time == 2019)
ggplot(data_2019, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point(alpha=0.7) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  scale_size_continuous(range = c(1, 10), breaks = c(1000000, 5000000, 10000000, 50000000, 100000000)) +
  xlab("GDP per capita") +
  ylab("Life expectancy") +
  labs(title = "GDP per capita versus life expectancy by country (2019)",
       subtitle = "Point size indicates total population, color indicates continent") +
  theme_bw()
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?
As we can see from the plots, the life expectancy and GDP per capita has improved across the entire world. However, Africa falls behind compared to other continent.
4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?

Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.
```{r}
avg_life_expectancy_1960 <- data %>%
  filter(time == 1960) %>% 
  group_by(region) %>% 
  summarize(avg_life_expectancy = mean(lifeExpectancy, na.rm = TRUE))
avg_life_expectancy_1960
```

```{r}
# Compute average life expectancy by continent in 2019
avg_life_expectancy_2019 <- data %>%
  filter(time == 2019) %>% 
  group_by(region) %>% 
  summarize(avg_life_expectancy = mean(lifeExpectancy, na.rm = TRUE))
avg_life_expectancy_2019
```

Yes it does 

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.

```{r}
data_growth <- data %>%
  group_by(region) %>%
  mutate(data_growth = (lifeExpectancy - 
  lag(lifeExpectancy))/lag(lifeExpectancy)) %>%
  filter(time > 1960 & time < 2019) %>%
  summarize(avg_growth = mean(data_growth, na.rm=TRUE)) %>%
  arrange(desc(avg_growth))
data_growth <- na.omit(data_growth)
data_growth
```

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
GDP_both <- data %>%
  filter(time %in% c(1960, 2019))

ggplot(GDP_both, aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge", bins = 50) +
  scale_fill_manual(values = c("purple", "orange"), name = "Year") +
  labs(title = "GDP per capita 1960 vs 2019", x = 
         "GDP per capita", y = "Frequency")
```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r}
data_with_name <- subset(data,!is.na(name))
rank_1960 <- data_with_name %>%
  filter(time == 1960) %>%
  arrange(desc(lifeExpectancy)) %>%
  mutate(rank = rank(-lifeExpectancy, na.last = TRUE)) %>% 
  filter(name == "United States of America")
rank_1960
```

```{r}
rank_2019 <- data_with_name %>%
  filter(time == 2019) %>%
  arrange(desc(lifeExpectancy)) %>% 
  mutate(rank = rank(-lifeExpectancy, na.last = "keep")) %>% 
  filter(name == "United States of America")
rank_2019
```
The US rank in 1960 is 17. The US rank in 2019 is 46.


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960
```{r}
n_countries_1960 <- data_with_name %>%
  filter(time == 1960 & !is.na(lifeExpectancy)) %>% 
  nrow()
rank_1960$rank/n_countries_1960
```
```{r}
n_countries_2019 <- data_with_name %>%
  filter(time == 2019 & !is.na(lifeExpectancy)) %>% 
  nrow()
rank_2019$rank/n_countries_2019
```

I spent 6 hours on this HW
  